Cryptographic Hash Functions

### 2019-04-12 math408/presentation/hashPaper.tex
Did work

A hash function is an algorithm that takes an input of arbitrary length and maps it through a series of transformations to a string of fixed length. The purpose of this is to verify a given input without needing to store that input.

@@

\tab There are several criteria that must be met for a hash function to be useful. They must be computationally efficient, which means the algorithm should run quickly on a computer without requiring a lot of memory or processing power.

They must have strong pre-image resistance, which means that given the output of a hash function, there should be no way to figure out the input that generated that particular output.

Also, hash functions have to be deterministic, meaning anytime a certain input is put into a hash function, the output should always be the same.

In order to be secure, hash functions must have strong collision resistance, which means there should be no reasonable way to find multiple inputs that generate the same output.

And finally, the output of a hash function should have the appearance of complete randomness.

@@

Password protection is one of the most critical uses of cryptographic hash functions. Consider a website that stores users' usernames and passwords in plain text. If a hacker gained access to the database, they would own the credentials to all users of that service. Since many people use the same password for all accounts, the attacker could pivot those credentials for such users to gain access to their emails, credit cards, and online banking accounts.

If instead a website stores the hashed version of users' passwords, an attacker could not reverse the hashes to discover the actual passwords of the users, and thus the hashes would be nearly useless to them.

@@

To get around this security issue, responsible websites use a technique called "salting". Using hash and salt, when a user creates an account and types a password in for the first time, the website uses a hashing algorithm, but also adds a random string of characters to the user-entered string, thus ensuring that even if hundreds of users enter the same exact password, each of their hashes will be different. Then every time the user enters their password to log in, the salt is added to whatever they entered and that result is hashed. If it matches the stored digest, they successfully login to their account.

This way, if a hacker gains access to a database, every hashed password will look different, and the only way they can match a password to a specific hash is by running a brute force attack.

@@

MD5 is still used extensively, but should no longer be implemented for any sensitive data verification, especially not for passwords or secure keys. Many common passwords can be found by typing their hash digests into Google search.

\section{SHA-256 (Secure Hashing Algorithm)}
\tab The current most commonly-used cryptographic hashing algorithm is the Secure Hashing Algorithm 2. There are several different algorithms in this family, but the most common one is the 256-bit SHA-256. Developed in 2001 by the NSA as part of the SHA-2 set, this is the most trusted and widely used hash algorithm in use currently. SHA-3 is the newest Secure Hashing Algorithm, and is slowly taking over and replacing SHA-256, but that will take time.

@@

Another word on collisions. Any well-developed hashing algorithm should have high collision resistance, but it is important to note that it is theoretically impossible to make a hash function that is completely collision-free. This is due to the fact that a hash function returns a fixed-length string. So, for a 128-bit hash, the output is always 32 characters. Possible characters include the 26 letters and 0-9, so a total of 36 possible characters. This gives $36^{32}$ possible hashes, which is $6.334 \times 10^{49}$. This is an astronomically high number, so breaking a hash of this size with brute force is infeasible, but given that there is a near-infinite number of inputs, there will always be possible collisions.

@@

	\bibitem{Tutorials Point Website}
	Tutorialspoint.com. (2019). Cryptography Hash functions. Retrieved from https://www.tutorialspoint.com/cryptography/cryptography\_hash\_functions.htm

>>>>

A hash function is an algorithm that takes an input of arbitrary length and maps it through a series of transformations to a string of fixed length. The purpose of this is to verify a given input without needing to store that input. The fixed length property is useful for storage and consistency, as well as file size reduction. It is important to understand what hash functions are and how they are used in everyday security so one can make good decisions about their personal digital protection.

@@

\tab There are several criteria that must be met for a hash function to be useful [1]. They must be computationally efficient, which means the algorithm should run quickly on a computer without requiring a lot of memory or processing power.

They must have strong pre-image resistance, which means that given the output of a hash function, there should be no way to figure out the input that generated that particular output. 
Also, hash functions have to be deterministic, meaning anytime a certain input is put into a hash function, the output should always be the same. 
In order to be secure, hash functions must have strong collision resistance, which means there should be no reasonable way to find multiple inputs that generate the same output. 
And finally, the output of a hash function should have the appearance of complete randomness.

@@

Password protection is one of the most critical uses of cryptographic hash functions. Consider a website that stores users' usernames and passwords in plain text. If a hacker gained access to the database, they would own the credentials to all users of that service. Since many people use the same password for all accounts, the attacker could pivot those credentials for such users to gain access to their emails, credit cards, and online banking accounts. 
If instead a website stores the hashed version of users' passwords, an attacker could not reverse the hashes to discover the actual passwords of the users, and thus the hashes would be nearly useless to them.

@@

To get around this security issue, responsible websites use a technique called "salting". Using hash and salt, when a user creates an account and types a password in for the first time, the website uses a hashing algorithm, but also adds a random string of characters to the user-entered string, thus ensuring that even if hundreds of users enter the same exact password, each of their hashes will be different. The salt is usually stored right next to the salted hash in the database. Then every time the user enters their password to log in, the salt is added to whatever they entered and that result is hashed. If it matches the stored digest, they successfully login to their account. 
This way, if a hacker gains access to a database, every hashed password will look different, and the only way they can match a password to a specific hash is by running a brute force attack.

@@

MD5 is still used extensively, but should no longer be implemented for any sensitive data verification, especially not for passwords or secure keys. [4]. Many common passwords can be found by typing their hash digests into Google search.

\section{SHA-256 (Secure Hashing Algorithm)}
\tab The current most commonly-used cryptographic hashing algorithm is the Secure Hashing Algorithm 2. There are several different algorithms in this family, but the most common one is the 256-bit SHA-256. Developed in 2001 by the NSA as part of the SHA-2 set, this is the most trusted and widely used hash algorithm in use currently. SHA-3 is the newest Secure Hashing Algorithm, and is slowly taking over and replacing SHA-256, but that will take time. [5].

@@

Another word on collisions. Any well-developed hashing algorithm should have high collision resistance, but it is important to note that it is theoretically impossible to make a hash function that is completely collision-free. This is due to the fact that a hash function returns a fixed-length string. So, for a 128-bit hash, the output is always 32 characters. Possible characters include the 26 letters and 0-9, so a total of 36 possible characters. This gives $36^{32}$ possible hashes, which is $6.334 \times 10^{49}$. This is an astronomically high number, so breaking a hash of this size with brute force is infeasible, but given that there is a near-infinite number of inputs, there will always be possible collisions. [2].

@@

	\bibitem{Tutorials Point Website}
	Tutorialspoint.com. (2019). Cryptography Hash functions. Retrieved from {https://www.tutorialspoint.com/cryptography/cryptography\_hash\_functions.htm}
